---
title: "Academic Innovation"
subtitle: "BA200 Fall 2019"
author: "Amy Lu, mingchlu@umich.edu"
date: "`r format.Date(Sys.Date(), '%B %d, %Y')`"
geometry: "left = 2cm, right = 2cm, top = 2cm, bottom = 2cm"
output: 
  html_document:
    code_folding: hide
    toc: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, echo = TRUE, result = "asis")

# Libraries: -------------------------------------------------------------------
library(tidyverse)
library(ggpubr)
library(DT)

# Read in data: ----------------------------------------------------------------
setwd("/Users/Amy/Desktop/AI/tandem/BA200")
BOT = read_csv("BA200_F19_BOT.csv")
EOT = read_csv("BA200_F19_EOT.csv")

# Data Cleaning: ---------------------------------------------------------------
colnames(BOT) = str_remove(colnames(BOT), "ba200__")
colnames(EOT) = str_remove(colnames(EOT), "ba200__")

# select key columns
vars_BOT = c("user_id", "sections", "Gender", "Location", "BT_Orientation", 
             "GroupPreference")
vars_EOT = c("user_id", "sections", "ET_Orientation", "ET_GroupPreference", 
             "Eval_Use", "Eval_Useful")

# Drop unused columns, drop rows with any missing values.
BOT = BOT %>% select(vars_BOT) %>% drop_na()
EOT = EOT %>% select(vars_EOT) %>% drop_na() %>% 
  filter(Eval_Use != -1 & Eval_Useful != -1)
```

## Objective
**Are there any significant differences in how students in different sections or with different demographics responded to BOT and EOT surveys?**

There are total of 616 students and 16 sections. The locations include Central, A2, North, Hill, and Far.

Variables interested:

1. `Eval_Use`:  
How interested would you be in using Tandem again? Scale from 1 to 5 (not at all interested to very interested).

2. `Eval_Useful`:  
How useful do you think Tandem was to you this term? Scale from 1 to 5 (not at all useful to very useful).

3. `BT/ET_Orientation`:  
If I could only get one outcome from BA200, it would be most important for me to: 
    - [Grade] Earn a good grade
    - [Performance] Be among the top performing teams in $Course
    - [Learn] Understand the subject well
    - [Practice] Get practice with the relevant skills
    - [Groupwork] Improve my ability to work with a group
    - [None] None of the above
    
4. `GroupPreference/ET_GroupPreference`:  
If given an option, I’d prefer to:
    - [Alone] Work alone
    - [Partner] Work with one partner
    - [Group] Work in a group

## Orientation

### Descriptive Statistics

```{r Descriptive - Orientation, fig.cap = cap, fig.align="center"}
BT_O = data.frame(table(BOT$sections, BOT$BT_Orientation))
colnames(BT_O) = c("sections", "Orientation", "BOT")
BT_O = BT_O %>% group_by(Orientation) %>% summarise(BOT = sum(BOT))

ET_O = data.frame(table(EOT$sections, EOT$ET_Orientation))
colnames(ET_O) = c("sections", "Orientation", "EOT")
ET_O = ET_O %>% group_by(Orientation) %>% summarise(EOT = sum(EOT))

df_O = left_join(BT_O, ET_O, by = c("Orientation"))

cap = paste0("**Figure 1.** *Number of students for each outcome.* Grade ",
            "remains the highest frequency while the frequency of Practice ",
            "decreases and the frequency of Groupwork increases as the ",
            "course moved toward the end of the term. The total counts are 616",
            " students for BOT and 530 students for EOT.")

pivot_longer(df_O, 2:3) %>%
  group_by(name) %>% arrange(-value) %>%
  mutate(Orientation = 
           factor(Orientation, 
                  levels = c("Grade", "Practice", "Performance", 
                             "Learn", "Groupwork", "None"))) %>%
  ggplot(aes(x = Orientation, y = value, fill = name)) +
  geom_bar(position="dodge", stat="identity") +
  geom_text(aes(label=value), vjust=-.8, color="black",
            position = position_dodge(0.9), size=3.5) +
  ylab("Counts") + labs(fill = "Survey") + xlab("") +
  scale_fill_manual(values = c("#FFcb05", "#00285E")) +
  theme_bw()
```

```{r BOT vs EOT Orientation}
t = as.table(cbind(as.numeric(df_O$BOT), as.numeric(df_O$EOT)))
dimnames(t) = list(Orientation = c("Grade", "Performance", "Learn", 
                                   "Practice", "Groupwork", "None"),
                   Term = c("BOT", "EOT"))
chisq = chisq.test(t, correct=FALSE)
```

*The Pearson's Chi-squared test shows that the `Orientation` for BOT and EOT are different at level $\alpha$ = .05 ($p$-value < 0.00).*

### Interaction between options
```{r Orientation - Interaction}
df = EOT %>% left_join(BOT, by = "user_id") %>%
  select(user_id, BT_Orientation, ET_Orientation) %>% drop_na()
df$BT_Orientation = factor(df$BT_Orientation,
                           levels = c("Grade", "Practice", "Performance", 
                                      "Learn", "Groupwork", "None"))
df$ET_Orientation = factor(df$ET_Orientation,
                           levels = c("Grade", "Practice", "Performance", 
                                      "Learn", "Groupwork", "None"))            

tbl = table(BOT = df$BT_Orientation, EOT = df$ET_Orientation)
tbl = prop.table(tbl, 1)

df_tbl = 
  as.data.frame(table(BOT = df$BT_Orientation, EOT = df$ET_Orientation)) %>%
  left_join(as.data.frame(tbl), by = c("BOT", "EOT")) %>%
  mutate(Freq = sprintf("%.0f (%.0f%%)", Freq.x, Freq.y*100)) %>%
  select(BOT, EOT, Freq) %>%
  rename(`Frequency (Percentage)` = Freq) %>%
  arrange(BOT)

wide = pivot_wider(df_tbl, names_from = EOT, values_from = `Frequency (Percentage)`)
names(wide)[1] = ""

knitr::kable(wide)
```

### Interaction between options by section
```{r Orientation - sections}
df_sec = EOT %>% left_join(BOT, by = c("user_id", "sections")) %>%
  transmute(user_id, BT_Orientation, ET_Orientation,
            sections = str_remove(sections, "Section 0"))
df_sec$BT_Orientation = factor(df_sec$BT_Orientation,
                           levels = c("Grade", "Practice", "Performance", 
                                      "Learn", "Groupwork", "None"))
df_sec$ET_Orientation = factor(df_sec$ET_Orientation,
                           levels = c("Grade", "Practice", "Performance", 
                                      "Learn", "Groupwork", "None"))  

tbl_sec = table(BOT = df_sec$BT_Orientation, EOT = df_sec$ET_Orientation, 
                Section = df_sec$sections)
tbl_sec = prop.table(tbl_sec, c(3,1))

df_tbl_sec = as.data.frame(prop.table(tbl_sec, c(3,1))) %>%
  transmute(Section, BOT, EOT, Freq = sprintf("%2.0f%%", Freq*100)) %>%
  arrange(BOT)

wide = pivot_wider(df_tbl_sec, names_from = EOT, values_from = Freq)
names(wide)[2] = "BOT \\ EOT"

options(DT.options = list(pageLength = 16))
datatable(wide, rownames = F, options = list(
  columnDefs = list(list(className = 'dt-center', targets = "_all"))))
```


```{r Orientation - Sections descriptive, fig.cap=cap, fig.width = 9}
BT_Os = BOT %>% group_by(sections, BT_Orientation) %>% 
  rename(Orientation = BT_Orientation) %>% summarise(BOT = n())
ET_Os = EOT %>% group_by(sections, ET_Orientation) %>% 
  rename(Orientation = ET_Orientation) %>% summarise(EOT = n())

df_Os = left_join(BT_Os, ET_Os, by = c("sections", "Orientation")) %>%
  ungroup() %>%
  mutate(sections = str_remove(sections, "Section 0"))

plot_line = df_Os %>% 
  mutate(Orientation = factor(Orientation, 
                              levels = c("Grade", "Practice", "Performance", 
                             "Learn", "Groupwork", "None"))) %>%
  group_by(Orientation) %>% 
  summarise(mB = mean(BOT), mE = mean(EOT, na.rm = T))

# Histogram
df_Os %>% pivot_longer(3:4, names_to = "Survey") %>%
  mutate(Orientation = factor(Orientation, 
                              levels = c("Grade", "Practice", "Performance", 
                                         "Learn", "Groupwork", "None"))) %>%
  ggplot(aes(x = sections, y = value, fill = Survey)) +
  geom_bar(na.rm = T, stat="identity", position="dodge") + 
  facet_wrap(~Orientation, nrow = 2) + ylab("Counts") +
  geom_hline(data = plot_line, aes(yintercept = mB), 
             color = "gold2", size = .5) +
  geom_hline(data = plot_line, aes(yintercept = mE), 
             color = "blue3", size = .5) +
  scale_fill_manual(values = c("#CFC096", "#00285E")) +
  theme_bw()

cap = paste0("**Figure 2.** *Number of students for each outcome of ",
             "Orientation for 16 sections.* The yellow and blue lines ",
             "represent the average counts for BOT and EOT respectively.")
```

## Group Preference

### Descriptive Statistics

```{r Descriptive - group preference, fig.align="center", fig.cap=cap}
BOT_gp = data.frame(table(BOT$sections, BOT$GroupPreference))
colnames(BOT_gp) = c("sections", "group_preference", "counts")
BT_gp = BOT_gp %>% group_by(group_preference) %>% summarise(BOT = sum(counts))

EOT_gp = data.frame(table(EOT$sections, EOT$ET_GroupPreference))
colnames(EOT_gp) = c("sections", "group_preference", "counts")
ET_gp = EOT_gp %>% group_by(group_preference) %>% summarise(EOT = sum(counts))

df_gp = left_join(BT_gp, ET_gp, by = c("group_preference"))

# Bar plot
pivot_longer(df_gp, 2:3) %>%
  group_by(name) %>% arrange(-value) %>%
  mutate(group_preference = 
           factor(group_preference, 
                  levels = c("Alone", "Partner", "Group"))) %>%
  ggplot(aes(x = group_preference, y = value, fill = name)) +
  geom_bar(position="dodge", stat="identity") +
  geom_text(aes(label=value), vjust=-.6, color="black",
            position = position_dodge(0.9), size=3.5) +
  ylab("Counts") + labs(fill = "Survey") + xlab("") +
  scale_fill_manual(values = c("#FFcb05", "#00285E")) +
  theme_bw()

cap = paste("**Figure 3.** *Number of students for each option.* Group ",
            "remained the highest frequency but the completion rate dropped ",
            "for the EOT survey.")
```

### Interaction between options
```{r Group Preference - Interaction}
df = EOT %>% left_join(BOT, by = "user_id") %>%
  select(user_id, GroupPreference, ET_GroupPreference) %>% drop_na() %>%
  mutate(BT_gp = factor(GroupPreference, 
                  levels = c("Alone", "Partner", "Group")),
         ET_gp = factor(ET_GroupPreference, 
                  levels = c("Alone", "Partner", "Group")))

tbl = table(BOT = df$BT_gp, EOT = df$ET_gp)
tbl = prop.table(tbl, 1)

df_tbl = 
  as.data.frame(table(BOT = df$BT_gp, EOT = df$ET_gp)) %>%
  left_join(as.data.frame(tbl), by = c("BOT", "EOT")) %>%
  mutate(Freq = sprintf("%.0f (%.0f%%)", Freq.x, Freq.y*100)) %>%
  select(BOT, EOT, Freq) %>%
  rename(`Frequency (Percentage)` = Freq) %>%
  arrange(BOT)

wide = pivot_wider(df_tbl, names_from = EOT, values_from = `Frequency (Percentage)`)
names(wide)[1] = ""

knitr::kable(wide)
```

### Interaction between options by section
```{r GP - sections}
df_sec = EOT %>% left_join(BOT, by = c("user_id", "sections")) %>%
  transmute(user_id, GroupPreference, ET_GroupPreference,
            sections = str_remove(sections, "Section 0")) %>% drop_na() %>%
  mutate(BOT = factor(GroupPreference, 
                  levels = c("Alone", "Partner", "Group")),
         EOT = factor(ET_GroupPreference, 
                  levels = c("Alone", "Partner", "Group")))

tbl_sec = table(BOT = df_sec$BOT, EOT = df_sec$EOT, Section = df_sec$sections)
tbl_sec = prop.table(tbl_sec, c(3,1))

df_tbl_sec = as.data.frame(prop.table(tbl_sec, c(3,1))) %>%
  transmute(Section, BOT, EOT, Freq = sprintf("%2.0f%%", Freq*100)) %>%
  arrange(BOT)

wide = pivot_wider(df_tbl_sec, names_from = EOT, values_from = Freq)
names(wide)[2] = "BOT \\ EOT"

options(DT.options = list(pageLength = 16))
datatable(wide, rownames = F, options = list(
  columnDefs = list(list(className = 'dt-center', targets = "_all"))))
```

```{r GP - Sections descriptive, fig.cap=cap, fig.width = 9}
BT_gps = BOT_gp %>% group_by(sections, group_preference) %>% 
  summarise(BOT = sum(counts))
ET_gps = EOT_gp %>% group_by(sections, group_preference) %>% 
  summarise(EOT = sum(counts))

df_gps = left_join(BT_gps, ET_gps, by = c("sections", "group_preference"))
        
# Average counts
plot_line = df_gps %>% 
  mutate(group_preference = 
           factor(group_preference, 
                  levels = c("Alone", "Partner", "Group"))) %>%
  group_by(group_preference) %>% 
  summarise(mB = mean(BOT), mE = mean(EOT, na.rm = T))

# Histogram
df_gps %>% pivot_longer(3:4, names_to = "Survey") %>%
  mutate(Sections = str_remove(sections, "Section 0"),
         group_preference = factor(group_preference, 
                                   levels = c("Alone", "Partner", "Group"))) %>%
  ggplot(aes(x = Sections, y = value, fill = Survey)) +
  geom_bar(na.rm = T, stat="identity", position="dodge") + 
  facet_wrap(~group_preference, nrow = 2) + 
  geom_hline(data = plot_line, aes(yintercept = mB), 
             color = "gold2", size = .5) +
  geom_hline(data = plot_line, aes(yintercept = mE), 
             color = "blue3", size = .5) +
  xlab("Sections") + ylab("Counts") + labs(fill = "Survey") +
  scale_fill_manual(values = c("#CFC096", "#00285E")) +
  theme_bw()

cap = paste0("**Figure 4.** *Number of students for each option of ",
             "GroupPreference for 16 sections.* The yellow and blue lines ",
             "represent the average counts for BOT and EOT respectively.")
```

## EOT - Eval_Use, Eval_Useful
### Eval_Use

```{r use table}
df_eval = left_join(EOT, BOT, by = c("user_id", "sections")) %>%
  filter(Gender == "Male" | Gender == "Female") %>%
  mutate(Sections = str_remove(sections, "Section 0"))

use = df_eval %>% group_by(Eval_Use, Gender) %>% 
  summarise(Counts = n()) %>%
  pivot_wider(names_from = Eval_Use, values_from = Counts) %>%
  rename(Eval_Use = Gender)

use$Eval_Use = as.character(use$Eval_Use)
use[3, ] = c("Total", colSums(use[,-1]))
knitr::kable(use)
```

```{r Eval_Use - all, fig.align="center", fig.cap=cap}
df_eval %>%
  ggplot(aes(x = Eval_Use, fill = Gender)) +
  geom_bar() +
  scale_fill_manual(values = c("#FFCB05", "#00285E"))
cap = paste0("**Figure 5.** *Number of students for Eval_Use (Scale 1 to 5 ",
             "represents not very interested to very interested.)* The plot ",
             "appears to be bimodal.")
```

```{r Eval_Use - section, fig.width=9, fig.align="center", fig.height=7, fig.cap=cap}
df_eval %>% group_by(sections) %>%
  ggplot(aes(x = Sections, fill = Gender)) +
  geom_bar() + 
  scale_fill_manual(values = c("#FFCB05", "#00285E")) +
  xlab("") +
  facet_wrap(~Eval_Use, nrow = 5) +
  theme_bw()
cap = paste0("**Figure 6.** *Number of students for each section.*")
```

### Eval_Useful

```{r useful table}
useful = df_eval %>% group_by(Eval_Useful, Gender) %>% 
  summarise(Counts = n()) %>%
  pivot_wider(names_from = Eval_Useful, values_from = Counts) %>%
  rename(Eval_Useful = Gender)

useful$Eval_Useful = as.character(useful$Eval_Useful)
useful[3, ] = c("Total", colSums(useful[,-1]))
knitr::kable(useful)
```

```{r Eval_Useful - all, fig.align="center", fig.cap=cap}
df_eval %>%
  ggplot(aes(x = Eval_Useful, fill = Gender)) +
  geom_bar() +
  scale_fill_manual(values = c("#FFCB05", "#00285E"))
cap = paste0("**Figure 7.** *Number of students for Eval_Useful (Scale 1 to 5 ",
             "represents not very useful to very useful).*")
```

```{r Eval_Useful - section, fig.width = 9, fig.align = "center", fig.height=7, fig.cap=cap}
df_eval %>% group_by(sections) %>%
  ggplot(aes(x = Sections, fill = Gender)) +
  geom_bar() + 
  scale_fill_manual(values = c("#FFCB05", "#00285E")) +
  xlab("") +
  facet_wrap(~Eval_Useful, nrow = 5) +
  theme_bw()
cap = paste0("**Figure 8.** *Number of students for each section.*")
```

